Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-01-17 paper/main.tex
Retest on NeoDTI data 1

\subsection{Datasets}
The data for the different parts of this model were obtained from various sources. Starting with the protein-protein interactions, we fetched $\approx11000$ human proteins with over 170.000 links from STRING (\cite{STRINGv10}). For the drug-target interactions themselves, we fetched 
137.000 links from STITCH database (\cite{STITCHv5}). As both STRING and STITCH provide probability scores for each association, we filtered them as advised by a threshold of 700, thus only obtaining likely interactions.\\
For the ontology segment we utilized PhenomeNET \citep{PhenomeNET2011}, a collection of various ontologies such as Human Phenotype Ontology \citep{HPO2018}, Gene Ontology (\citet{GOoriginal2000} \citet{GOrecent2020}), Mammalian Phenotype Ontology \citep{MP2009} and numerous others. \todo{which ontologies to cite}
Side effects and their links to drugs were obtained Side Effect Resource (SIDER)\citep{SIDER} and structured according MedDRA database \citep{MedDRA}. They were mapped to PhenomeNET with aid of \textit{Phenomebrowser.net}, which provides a SPARQL query endpoint for the mentioned resources. \\

We additionally obtained molecular structure based features for drugs from \textit{SmilesTransformer}\citep{SmilesTransformer} and proteins from \textit{DeepGOPlus}\citep{DeepGoPlus}.\\

@@

\subsection{Model}
In order to build a method that incorporates both top-down and bottom-up features, we first created a model for each. In the top-down section, we used \textit{DL2vec}\citep{DL2vec2020} to obtain ontology based representations. Hereby, DL2vec constructs a graph by introducing vertices and edges for each ontology class and axiom, respectively, followed by random walks starting from each entity. These walks are eventually learned on using a Word2vec \citep{Word2vec2013} model. Thus, we pick up rich, neighbourhood focused representations for each entity, which has shown great results for representing protein function and phenotypes. The overall structure of the ontology can be seen in Figure~\ref{fig:Onto}.\\

@@

\subsection{High level description}
To obtain a prediction for a given drug-protein pair, we first need to pretrain the siamese networks for both molecular and ontology features, yielding the deep neural network transformers for both methods and both drugs and proteins, individually. These embeddings are now added to the graph neural network as node features, where we start an end-to-end learning and training process. Notably, like in the pretraining phase, we rely on siamese neural networks, enforcing similar representations for fitting drug-protein pairs. The overall architecture can be seen in Figure~\ref{fig:Fullmodel}.

>>>>

\subsection{Datasets}
The data for the different parts of this model were obtained from various sources. Starting with the protein-protein interactions, we fetched $\approx11000$ human proteins with over 170.000 links from STRING \citep{STRINGv10}. For the drug-target interactions themselves, we fetched 
137.000 links from STITCH database \citep{STITCHv5}. As both STRING and STITCH provide probability scores for each association, we filtered them as advised by a threshold of 700, thus only obtaining likely interactions.\\
For the ontology segment we utilized PhenomeNET \citep{PhenomeNET2011}, a collection of various ontologies such as Human Phenotype Ontology \citep{HPO2018}, Gene Ontology (\citet{GOoriginal2000} \citet{GOrecent2020}), Mammalian Phenotype Ontology \citep{MP2009} and numerous others. \todo{which ontologies to cite}
Side effects and their links to drugs were obtained Side Effect Resource (SIDER)\citep{SIDER} and structured according MedDRA database \citep{MedDRA}. They were mapped to PhenomeNET with aid of \textit{Phenomebrowser.net}, which provides a SPARQL query endpoint for the mentioned resources. \\

@@

\subsection{Model}
In order to build a method that incorporates both top-down and bottom-up features, we first created a model for each individually. Hence, we assemble rich molecular structure based features for drugs from \textit{SmilesTransformer}\citep{SmilesTransformer} and proteins from \textit{DeepGOPlus}\citep{DeepGoPlus}. \textit{SmilesTransformer} introduces an autoencoder, learning over the SMILES strings and thus the molecular organization of each drug in an unsupervised manner. On the other hand, we take advantage of the pretrained models of \textit{DeepGOPlus}, obtaining features from the proteins amino acid sequence, showing significant performance in the field of protein function prediction. Thus, both embeddings seem to suitably supplement the  following ontology based represenations.\\

In the top-down section, we used \textit{DL2vec}\citep{DL2vec2020} to obtain ontology based representations. Hereby, DL2vec constructs a graph by introducing vertices and edges for each ontology class and axiom, respectively, followed by random walks starting from each entity. These walks are eventually learned on using a Word2vec \citep{Word2vec2013} model. Thus, we pick up rich, neighbourhood focused representations for each entity, which has shown great results for representing protein function and phenotypes. The overall structure of the ontology can be seen in Figure~\ref{fig:Onto}.\\

@@

\subsection{High level description}
To obtain a prediction for a given drug-protein pair, we first build DL2vec representations for both drugs and proteins over PhenomeNET, while also preparing structural, molecular representations. Hence, we pretrain the siamese networks for both molecular and ontology features, yielding the deep neural network transformers for both methods and both drugs and proteins, individually. These embeddings are now added to the graph neural network as node features, where we start an end-to-end learning and training process. Notably, like in the pretraining phase, we rely on siamese neural networks, enforcing similar representations for fitting drug-protein pairs. The overall architecture can be seen in Figure~\ref{fig:Fullmodel}.
